Overlapping snow algae communities in the alpine and subalpine zones of British Columbia

### 2019-12-28 manuscript/manuscript.Rmd
re wrote intro

---
title: "Contrasting snow algae communities in the mountains of British Columbia, Canada"
# who grows in the snow? multi-hued snow algae communities
# different coloured alpine and subalpine snows harbour constrasting snow algae communities
# contrasting snow algae communities in di
# not just dirty snow; alpine snow algae communities have different assemblages
# green, orange, or red snow caused by distinct
# snow algae communities causing colored snow in alpine and subalpine summer snow 
# distinct snow algae communities across an elevational gradient
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
csl: templates/nature-communications.csl
bibliography: casey1.bib
abstract: |
  Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt. Green, orange, and red patches of snow algae can harbour morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood. We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada. Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups of *Chlainomonas* and several potentially novel clades. Above tree line communities were dominated by *Sanguina*, and below tree line by *Chloromonas*. *Chlainomonas* was abundant in samples across all elevations. We also observed blooms of *Chloromonas krienitzii* transition from green to orange throughout the season. We found snow algae community composition was highly variable from patch to patch, even on the same mountain. Although relatively few genera dominated snow algae communities, we found high diversity at the species level. 
---

@@

# Introduction {-}

Snow algae are cold-adapted microalgae that bloom on melting summer snow throughout alpine and polar regions. Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017]. Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019].

Sanger seq isolates

Sanger seq field samples

18S HTS

<!-- re work this to follow the flow of my presentation -->
Microscopy reveals a diversity of algae in patches of green, orange, or pink snow (Fig. 1). Amplicon high-throughput sequencing (HTS) shows that most snow algae are typically Chlamydomonadales, although other orders are occasionally reported[@remias_two_2019;@lutz_integrated_2015;@soto_microbial_2020]. Three genera are frequently found in colored snow. *Sanguina* are red or orange cells found across the Arctic, which was recently re-assigned to its own genus based on Sanger sequencing of field samples[@prochazkova_sanguina_2019].  *Chloromonas* currently has 35 species listed on GenBank, 13 of which are snow algae[@matsuzaki_taxonomic_2019]. Microscopy suggests *Chloromonas* is prevalent in forested sites[@hoham_distribution_1979,@nedbalova_ecology_2008], but recent 18S metabarcoding found *Chloromonas* was also widespread in Arctic red snow in low relative abundance[@lutz_biogeography_2016]. The third genus *Chlainomonas* looks completely different from *Chloromonas*, but *rbcL* phylogeny suggests the two are closely related[@novis_two_2008]. Both *Chlainomonas* and *Sanguina* are colored red by astaxanthin, but *Chlainomonas* cells are typically twice the diameter, with four flagella and a distinct cell wall[@remias_ecophysiology_2016] (Fig 2x). *Chlainomonas* and *Sanguina* remain uncultured[@prochazkova_ecophysiological_2018;@novis_two_2008].

Microscopic identification is complicated by the multiple life stages typical of microalgae: the same species can look completely different between life stages; conversely, the same species can look nearly identical. The freshwater algae *Haematococcus pluvialis* undergoes mitosis as green cells, but stressors such as high light and nutrient starvation triggers formation of a dormant cyst with thick cell walls and astaxanthin pigment[@ota_carotenoid_2018;@kobayashi_light-independent_1997]. Attempts to induce cyst development in snow algae remain unsuccessful; the bright pigments and ornamented cell walls of *Chloromonas* observed in the field bear little resemblance to the green cells in culture[@leya_response_2009]. Amplicon studies have found that green snow is caused by different species than red snow[@lutz_integrated_2015;@terashima_microbial_2017;@soto_microbial_2020], so it remains unknown whether red snow species undergo a green life phase at all.

Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015]. Despite nearly a century of microscopic study of red snow[@leya_snow_2013], how snow algae blooms develop remains a mystery. How do cells propagate in this harsh environment? What limits the extent of snow algae blooms? Critically, we lack basic understanding of snow algae diversity and life history.

<!-- this from draft 1 of casey1, above is from MBB806 -->
 Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsen drought [@Barnhart2016]. Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form. We don't know how cells colonize the white snow to initiate the bloom, what triggers snow algae to divide and produce pigment, or why one patch of snow turns red while the neighboring snow remains white. Critically, we lack basic understanding of the species composition and diversity present in snow algae blooms, which is fundamental to resolving long-standing mysteries about snow algae bloom formation.

Snow algae blooms can color snow red, orange, or green, and can one or contain multiple species of algae. Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophyceae [@Remias2013;@Remias2019] have been reported. Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin responsible for the red color of some snow algae blooms. Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019a;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975b]. However, some species are known from only one or two samples, and the recently described *Chlainomonas* and *Sanguina* are poorly explored, with only two species in each.

Snow algae community composition is largely unknown. Some studies batch sequencing samples, and microscopy suggests that blooms are dominated by a single species. The red cells of *Sanguina nivaloides* are widespread and cosmopolitan [@Prochazkova2019a;@Segawa2018], but 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae (likely *Sanguina*, as it had not been described at the time), these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites. 33 red snow samples from across the Arctic were relatively homogeneous in community composition [@Lutz2016a], but genetic comparison of the morphologically diverse snow algae communities found in mountains from temperate latitudes [@Nedbalova2008a] is lacking.

Despite over a century of study, no one has successfully cultured the snow algae that produce red secondary pigments. In model organism green algae *Chlamydomonas reinhardtii* cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Salome2019], thus red snow algae could originate from green cells. Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow. Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashima2017]; however it remains an open question whether green snow can develop into red snow.

In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy and 18S and *rbc*L amplicon HTS. The *rbc*L gene (coding for the large subunit of Rubisco) specifically targets phototrophs, and is more differentiated than the traditional 18S ribosomal marker gene, thus allowing us to specifically target snow algae. Our goal was to see what species were present, and how community composition varies over elevation.

@@

# Conclusions

@@

We used rbcL as a marker gene, which offers high resolution between microalgae species [@Zou2016], and has reference data for snow algae available on GenBank. We also sequenced each sample with 18S primers, due to it's coverage across a wide range of taxa, and its ubiquity in reference databases. We designed rbcL primers to target an approximately 400 bp section of this gene, based on 20 snow algae rbcL GenBank sequences from *Chloromonas* and *Chlainomonas* (GenBank accession numbers AB434272.1, LC012752.1, LC012747.1, AF517072.1, LC012738.1, LC012739.1, AB434267.1, EU030690.1, LC360494.1, AJ001878.1, AB022225.1, DQ885964.2, DQ885962.1, AJ001879.1, AB022226.1, AB022530.1, LC012751.1, AB504764.1, EU030689.1, AB101508.1). *Sanguina* sequences were not included because they were not available at the time. We designed primers using the primer design tool from the Eurofins Genomics website (www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/). The resultant primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’). To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3') [@Wang2014].

>>>>

---
title: "Overlapping snow algae communities in the alpine and subalpine zones of British Columbia"
# who grows in the snow? multi-hued snow algae communities
# different coloured alpine and subalpine snows harbour constrasting snow algae communities
# contrasting snow algae communities in di
# not just dirty snow; alpine snow algae communities have different assemblages
# green, orange, or red snow caused by distinct
# snow algae communities causing colored snow in alpine and subalpine summer snow 
# distinct snow algae communities across an elevational gradient
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
csl: templates/nature-communications.csl
bibliography: casey1.bib
abstract: |
  Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt. Green, orange, and red patches of snow algae can harbour morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood. We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada. Communities were dominated by algae closely related to *Sanguina*, *Chloromonas*, and *Chlainomonas*, as well as potentially novel taxa. Above treeline communities were dominated by *Sanguina*, and below tree line by *Chloromonas*. *Chlainomonas* was abundant in samples across all elevations. We also observed blooms of *Chloromonas krienitzii* transition from green to orange throughout the season. We found snow algae community composition was highly variable from patch to patch, even on the same mountain. Although relatively few genera dominated snow algae communities, we found high diversity at the species level. 
---

@@

# Introduction {-}
<!-- audience, problem -->
Reduction in cryospheric albedo has become the focus of attention in recent years as an important feedback loop of global warming. Microbial growth on snow surface can reduce albedo, notably when snow algae form extensive red blooms on melting summer snow[@ganey_role_2017;@lutz_biogeography_2016;@onuma_temporal_2019]. Snow algae blooms are found on every continent, and on snow overlying sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996]. Despite their global ubiquity, we are only beginning to identify the various types of algae communities that cause snow algae blooms. 
<!-- why do we care if there are different communities as a climate scientist? maybe this should be a paper for snow phycologists? -->
<!-- different communities could have different ecology and life history -->
<!-- we need to define the community types to understand ecology of different blooms.  -->

<!-- snow algae taxonomy -->
Most snow algae are green algae (Chlorophyceae), although recent studies suggest that other families of algae can also cause snow algae blooms[@lutz_omics_2015;@soto_microbial_2020;@remias_two_2019]. Within the Chlorophyceae *Sanguina nivaloides* is prevalent in red snow worldwide, but red snow also contains many other clades which have yet to be named[@prochazkova_sanguina_2019]. Other Chlorophyceaen snow algae include *Chloromonas* and the closely-related *Chlainomonas*. Currently, *Sanguina* contains two species, *Chloromonas* fourteen, and *Chlainomonas* two[@matsuzaki_taxonomy_2019;@novis_two_2008]. Only *Chloromonas* has been sucessfully cultivated in the lab.

<!-- microscopy suggests distinct communities -->
Microscopy suggests that snow algae blooms are in many cases are dominated by single species. In the mountains of southwestern USA and Czech Republic red snow was prevelant above treeline, dominated by cells similar in appearance to *S. nivaloides*, while below treeline green or orange snow were prevalent dominated by cells similar to *Chloromonas*[@hoham_distribution_1979;nedbalova_ecology_2008]. Red snow dominated by *Chlainomonas* type cells were reported from slushy snow overlying alpine lakes[@novis_two_2008;@remias_ecophysiology_2016]. However, microscopic identification of snow algae is complicated by the fact that the same species can look completely different at different life stages. Isolates of *Chloromonas krienitzii* are green, bean-shaped biflagellates, but genetically identical isolated single-cells from field samples appear as orange spheres with thick cell walls, short spines, and no flagella[@matsuzaki_revisiting_2015]. Conversely, different species can look nearly identical: recent phylogenetic work shows that four similar looking cells previously referred to as *Chloromonas cf. nivalis* actually contain multiple genetically distinct clades[@matsuzaki_taxonomic_2019].

<!-- but genetic comparison, most look the same? -->
Although snow algae communities contain microscopically distinct snow algae community types, studies using 18S amplicon high-throughput sequencing (metabarcoding) suggests that snow algae communities are highly similar. 33 red snow communities across the Arctic were remarkably similar: all were dominated by uncultured Chlamydomonadaceae, with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale*[@lutz_biogeography_2016]. Another study using 18S metabarcoding of snow algae in the mountains of Japan found neighboring red and green snow patches with  highly overlapping communities, although the red snow snow contained one OTU not found in the green communities[@terashima_microbial_2017]. However, side-by-side red and green snow in Greenland contained completely different communities, and the authors concluded that red and green snow were caused by a different species[@lutz_omics_2015].

Given how little is known about snow algae distribution and diversity, particularly in alpine habitats, our goal is to characterise the snow algae community structure in alpine and subalpine habitats. We surveyed snow algae communities from throughout the Coast Range of British Columbia, Canada using light microscopy and metabarcoding. 18S is highly conserved, making it an ideal universal barcode for eukaryotes, but at lower taxonomic levels 18S is too similar to resolve closely related taxa. To complement our 18S sequencing, we also targeted the more differentiated *rbcL* coding for rubisco. By using three cross-referenced metrics of relative abundance, we were able to describe the community structure with high taxonomic resolution. We found communities above treeline contained a distinct set of taxa, and instead of sharply defined community types we found something more akin to a gradient, with varying relative abundances of different snow algae taxa.

<!--  This is an important first step before we measure habitat variables, will allow us to choose which variables to measure... -->
<!-- significance. different community, different ecology and or life history. and the snow algae are the primary producers of the community. define the study system.  -->
<!-- hypothesis: there are distinct communities of snow algae-->
<!-- methods: cell count, 18s, and rbcL amplicon sequencing -->
<!-- results: there is much overlap between communities, but many with dominant species. Some chloromonas, raphi only found with sanguina. Chlainomonas widespread. overall, no, there are not sharply delineated communities. -->
<!-- new questions raised: what habitat differences are found with different communities? is bass becking applicable here? what ecological differences, life history differences characterise the different communities?  -->

@@

 Hypothetically, a positive feedback loop could permit snow algae blooms to propogate, in which snow algae melt snow, producing more liquid water that permits more snow algae growth[].

@@

We used rbcL as a marker gene, which offers high resolution between microalgae species [@Zou2016], and has reference data for snow algae available on GenBank. We also sequenced each sample with 18S primers, due to it's coverage across a wide range of taxa, and its ubiquity in reference databases. We designed rbcL primers to target an approximately 400 bp section of this gene, based on 20 snow algae rbcL GenBank sequences from *Chloromonas* and *Chlainomonas* (GenBank accession numbers AB434272.1, LC012752.1, LC012747.1, AF517072.1, LC012738.1, LC012739.1, AB434267.1, EU030690.1, LC360494.1, AJ001878.1, AB022225.1, DQ885964.2, DQ885962.1, AJ001879.1, AB022226.1, AB022530.1, LC012751.1, AB504764.1, EU030689.1, AB101508.1). *Sanguina* sequences were not included because they were not available at the time. We designed primers using the primer design tool from the Eurofins Genomics website (www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/). The resultant primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’). To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3'), targeting the V7-V8 hypervariable regions[@Wang2014;@].
